\ab{Constraints on quasi--normal-mode frequencies with LIGO-Virgo binary--black-hole observations}

### 2021-03-25 papers/intro_paper/intro.tex
committed all remaining changes

The observation of GWs has had significant astrophysical and cosmological
implications~\cite{TheLIGOScientific:2016htt,GBM:2017lvd,Monitor:2017mdv,Abbott:2017xzu}. It
has also allowed us to probe fundamental
physics and test predictions of Einstein's theory of General Relativity
(GR) in the previously unexplored highly-dynamical, and strong field
regime~\cite{TheLIGOScientific:2016src,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks}. In GR, a binary black hole (BBH) system is described by
three distinct phases: an early \textit{inspiral}~\cite{Blanchet:2013haa}, where the two
compact objects spiral in loosing energy because of the emission of GWs, a \textit{merger}~\cite{Pretorius:2005gq,Campanelli:2005dd,Baker:2005vv} marked by the
formation of a common apparent horizon, and a \textit{ringdown}~\cite{Vishveshwara:1970zz,Vishveshwara:1970cc,Press:1971wr,Chandrasekhar:1975zza,Detweiler:1980gk}, during which the newly formed remnant object settles down to a Kerr BH emitting quasi-normal-modes (QNMs) (i.e., damped oscillations with specific, discrete frequencies and decay times). \comment{AB: I have left only  the QNM refs until 80s, and added the main refs for ``inspiral'' and ``merger''. I found the citation unbalanced.}

@@

The first paper on tests of GR by the LIGO Collaboration~\cite{TheLIGOScientific:2016src} also provided us with the first measurement of the dominant damped-oscillation signal in the ringdown stage of a BBH
coalescence, and more recently a similar measurement was made with the high-mass event GW190521~\cite{Abbott:2020tfl,Abbott:2020mjq}. The set of available measurements was greatly expanded in the latest LIGO-Virgo O3a testing GR paper~\cite{Abbott:2020jks} where a comprehensive analysis of the properties of the remnant object, including the ringdown stage, was reported for tens of GW events. The no-hair conjecture in GR~\cite{Carter:1971zc,Hansen:1974zz} states that an (electrically neutral) astrophysical BH is completely described by two observables: mass and spin angular momentum \ab{(for short spin)}. One
consequence of the no-hair conjecture is that the (complex) QNM
frequencies of gravitational radiation emitted by a perturbed isolated
BH are uniquely determined by its mass and spin. Hence
a test of the no-hair conjecture would involve checking for
consistency between estimates of mass and spin of the remnant object
across multiple QNM frequencies~\cite{Dreyer:2003bv,Berti:2005ys}. An inconsistency would either
indicate a non-BH nature of the remnant object, or an incompleteness
of GR as the underlying theory of gravity.

The consistency between the post-merger signal and the least damped QNM was first demonstrated in
Ref.~\cite{TheLIGOScientific:2016src} for GW150914, and confirmed through several independent analyses~\cite{Brito:2018rfr,Carullo:2019flw,Isi:2019aib,CalderonBustillo:2020tjf}. This was later extended to include
overtones in Refs.~\cite{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw,Abbott:2020jks}. Consistency of the late-time waveform with a single QNM is a test of the ringdown of
a BBH coalescence, but not necessarily a test of the no-hair
conjecture, which requires the measurement of (at least) two QNMs (BH
spectroscopy), and consistency between \ab{them~\cite{Dreyer:2003bv,Carullo:2018sfu,Carullo:2018gah,Bhagwat:2019bwv,Cabero:2019zyt,Maselli:2019mjd}}. The
nature of the remnant object has also been explored through tests of
BH thermodynamics, like the Hawking's area
theorem~\cite{Cabero:2017avf,Isi:2020tac} or through search for echos in the
post-merger signal~\cite{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}.  None of these tests have found evidence for non-BH nature of the remnant
object (as described in GR) in LIGO-Virgo BBH observations.

@@

The rest of the paper is organized as follows. Section~\ref{sec:model} describes our parameterized IMR
waveform model. In Sec.~\ref{sec:method}, we define our framework to test GR, notably how we can measure
complex QNM frequencies with our parameterized model within a Bayesian formalism\ab{, and validate it using 
several synthetic-signal injection studies in Gaussian noise}. Then, in Sec.~\ref{sec:results},
we \sout{demonstrate} \ab{apply}  our method on real GW events \ab{using LIGO and Virgo data}.\sout{, as well as simulated signals}. In particular, we analyze
the GWTC 1/2  events and \ab{obtain the most stringent constraints on the dominant (complex)} QNM frequecies \ab{with our method}\sout{to be \macro{$\df{220} = 0.02^{+0.03}_{-0.03}$} and \macro{$\dtau{220} = 0.11 ^{+0.12} _{-0.12}$}}. \comment{AB: I think it is too early to give the results here.} Finally, in Sec.~\ref{sec:discussion} we provide a summary of
our results and discuss future developments.

>>>>

The observation of GWs has had significant astrophysical and cosmological
implications~\cite{TheLIGOScientific:2016htt,GBM:2017lvd,Monitor:2017mdv,Abbott:2017xzu}. It
has also allowed us to probe fundamental
physics and test predictions of Einstein's theory of General Relativity
(GR) in the previously unexplored highly-dynamical, and strong field
regime~\cite{TheLIGOScientific:2016src,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks}. In GR, a binary black hole (BBH) system is described by
three distinct phases: an early \textit{inspiral}~\cite{Blanchet:2013haa}, where the two
compact objects spiral in loosing energy because of the emission of GWs, a \textit{merger}~\cite{Pretorius:2005gq,Campanelli:2005dd,Baker:2005vv} marked by the
formation of a common apparent horizon, and a \textit{ringdown}~\cite{Vishveshwara:1970zz,Vishveshwara:1970cc,Press:1971wr,Chandrasekhar:1975zza,Detweiler:1980gk}, during which the newly formed remnant object settles down to a Kerr BH emitting quasi-normal-modes (QNMs) (i.e., damped oscillations with specific, discrete frequencies and decay times).

@@

The first paper on tests of GR by the LIGO Collaboration~\cite{TheLIGOScientific:2016src} also provided us with the first measurement of the dominant damped-oscillation signal in the ringdown stage of a BBH
coalescence, and more recently a similar measurement was made with the high-mass event GW190521~\cite{Abbott:2020tfl,Abbott:2020mjq}. The set of available measurements was greatly expanded in the latest LIGO-Virgo O3a testing GR paper~\cite{Abbott:2020jks} where a comprehensive analysis of the properties of the remnant object, including the ringdown stage, was reported for tens of GW events. The no-hair conjecture in GR~\cite{Carter:1971zc,Hansen:1974zz} states that an (electrically neutral) astrophysical BH is completely described by two observables: mass and spin angular momentum \ab{(for short spin)}\abhi{small?}. One
consequence of the no-hair conjecture is that the (complex) QNM
frequencies of gravitational radiation emitted by a perturbed isolated
BH are uniquely determined by its mass and spin. Hence
a test of the no-hair conjecture would involve checking for
consistency between estimates of mass and spin of the remnant object
across multiple QNM frequencies~\cite{Dreyer:2003bv,Berti:2005ys}. An inconsistency would either
indicate a non-BH nature of the remnant object, or an incompleteness
of GR as the underlying theory of gravity.

The consistency between the post-merger signal and the least damped QNM was first demonstrated in
Ref.~\cite{TheLIGOScientific:2016src} for GW150914, and confirmed through several independent analyses~\cite{Brito:2018rfr,Carullo:2019flw,Isi:2019aib,CalderonBustillo:2020tjf}. This was later extended to include
overtones in Refs.~\cite{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw,Abbott:2020jks}. Consistency of the late-time waveform with a single QNM is a test of the ringdown of
a BBH coalescence, but not necessarily a test of the no-hair
conjecture, which requires the measurement of (at least) two QNMs (BH
spectroscopy), and consistency between them~\cite{Dreyer:2003bv,Carullo:2018sfu,Carullo:2018gah,Bhagwat:2019bwv,Cabero:2019zyt,Maselli:2019mjd}. The
nature of the remnant object has also been explored through tests of
BH thermodynamics, like the Hawking's area
theorem~\cite{Cabero:2017avf,Isi:2020tac} or through search for echos in the
post-merger signal~\cite{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}.  None of these tests have found evidence for non-BH nature of the remnant
object (as described in GR) in LIGO-Virgo BBH observations.

@@

The rest of the paper is organized as follows. Section~\ref{sec:model} describes our parameterized IMR
waveform model. In Sec.~\ref{sec:method}, we define our framework to test GR, notably how we can measure
complex QNM frequencies with our parameterized model within a Bayesian formalism, and validate it using 
several synthetic-signal injection studies in Gaussian noise. Then, in Sec.~\ref{sec:results},
we apply our method on real GW events using LIGO and Virgo data. In particular, we analyze
the GWTC 1/2  events and obtain the most stringent constraints on the dominant (complex) QNM frequecies. Finally, in Sec.~\ref{sec:discussion} we provide a summary of
our results and discuss future developments.

### 2021-03-25 papers/intro_paper/model_method.tex
committed all remaining changes

A GW signal from the (quasi-circular) coalescence of two BHs is
completely described in GR by $15$ parameters,
$\bxigr$. These can be grouped into the \emph{intrinsic} parameters:
the masses, $m_1, m_2$ and spins, $\bs1, \bs2$ of the component
objects in the binary; and the
\emph{extrinsic} parameters: a reference time $t_c$ and phase
$\phi_c$, the sky position of the binary ($\alpha$,
$\delta$), the luminosity distance, $d_L$, and the binary's orientation
described through the inclination of the binary $\iota$ and its
polarization $\psi$. We also introduce the total mass $M = m_1+m_2$,
and the (dimensionless) symmetric mass ratio $\nu = m_1m_2/M^2$.

@@

In the $\SEOB$ model constructed in Ref.~\cite{Cotesta:2018fcv}, the
complex frequencies $\sigma_{\ell m 0}$ are expressed in terms of the
final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the
latter are related to the BBH's component masses and spins through
NR--fitting-formulas obtained in
GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the
spirit of what was done in \paperone, we promote the QNM (complex)
frequencies to be free parameters of the model, while keeping the
inspiral-plunge modes $h_{\ell m}^\mathrm{inspiral-plunge}(t)$ fixed
to their GR values. More explicitly, we introduce a parameterized
version of the $\SEOB$ model where the frequency and the
damping time of the ${\ell m 0}$ mode (i.e, $(f_{\ell m 0}, \tau
_{\ell m 0})$) are defined through the fractional deviations, $(\delta
f_{\ell m 0},\delta \tau_{\ell m 0})$, from the corresponding GR
values.\sout{, which are obtained using NR fits}
Thus~\cite{Gossan:2011ha,Meidam:2014jpa},
\begin{subequations}
\begin{eqnarray}
f_{\ell m 0} &=& f_{\ell m 0}^{\text{GR}}\, (1 + \delta f_{\ell m 0})\,,\label{eq:nongr_freqs_a} \\ 
\tau _{\ell m 0} &=& \tau _{\ell m 0}^{\text{GR}}\, (1 + \delta \tau_{\ell m 0})\,. \label{eq:nongr_freqs_b}
\end{eqnarray}
\end{subequations}
We denote such a parameterized waveform model $\pSEOB$~\footnote{This
waveform model is called {\tt pSEOBNRv4HM} in LAL.}.

\sout{As said,} The GR quantities $( f_{\ell m 0}^{GR},\tau_{\ell m 0}^{GR})$ are
constructed using the \sout{same} NR--fitting--formula \ab{from Refs.~\cite{Taracchini:2013rva,Hofmann:2016yih}}.
\sout{described above.} We note that when leaving $\sigma_{\ell m}$ to vary
freely, the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell
  m}(t)$ in general also differ from the GR predictions, since
those functions depend on the QNM complex frequencies, as can be seen
from the expressions for $c_{i,c}^{\ell m}$ and $d_{1,c}^{\ell m}$ in Eqs.~(\ref{c1}),
(\ref{c2}), and (\ref{d1}). \ab{As a consequence, the ringdown signal (amplitude and phase) 
soon after merger deviates from the one predicted by GR.}

>>>>

A GW signal from the (quasi-circular) coalescence of two BHs is
completely described in GR by $15$ parameters,
$\bxigr$. These can be grouped into the \emph{intrinsic} parameters:
the (source) masses, $m_1, m_2$ and spins, $\bs1, \bs2$ of the component
objects in the binary; and the
\emph{extrinsic} parameters: a reference time $t_c$ and phase
$\phi_c$, the sky position of the binary ($\alpha$,
$\delta$), the luminosity distance, $d_L$, and the binary's orientation
described through the inclination of the binary $\iota$ and its
polarization $\psi$. We also introduce the total (source) mass $M = m_1+m_2$,
and the (dimensionless) symmetric mass ratio $\nu = m_1m_2/M^2$. When needed, 
we shall denote the detector masses as $m_{1\,\rm det}= (1+z)\,m_1$ and $m_{2 \,\rm det} = (1+z)\,m_2$, 
where $z$ is the redshift. \abhi{We also follow the convention $m_1 > m_2$, and hence the asymmetric mass ratio, $q = m_1/m_2 \geq1$.}

@@

In the $\SEOB$ model constructed in Ref.~\cite{Cotesta:2018fcv}, the
complex frequencies $\sigma_{\ell m 0}$ are expressed in terms of the
final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the
latter are related to the BBH's component masses and spins through
NR--fitting-formulas obtained in
GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the
spirit of what was done in \paperone, we promote the QNM (complex)
frequencies to be free parameters of the model, while keeping the
inspiral-plunge modes $h_{\ell m}^\mathrm{inspiral-plunge}(t)$ fixed
to their GR values. More explicitly, we introduce a parameterized
version of the $\SEOB$ model where the frequency and the
damping time of the ${\ell m 0}$ mode (i.e, $(f_{\ell m 0}, \tau
_{\ell m 0})$) are defined through the fractional deviations, $(\delta
f_{\ell m 0},\delta \tau_{\ell m 0})$, from the corresponding GR
values~\cite{Gossan:2011ha,Meidam:2014jpa}.

Thus,
\begin{subequations}
\begin{eqnarray}
f_{\ell m 0} &=& f_{\ell m 0}^{\text{GR}}\, (1 + \delta f_{\ell m 0})\,,\label{eq:nongr_freqs_a} \\ 
\tau _{\ell m 0} &=& \tau _{\ell m 0}^{\text{GR}}\, (1 + \delta \tau_{\ell m 0})\,. \label{eq:nongr_freqs_b}
\end{eqnarray}
\end{subequations}

\comment{AB: as discussed, please expalin very clearly how the GR frequency and decay times are computed, and mark in the equation above 
their dependence on the mass and spin parameters.}\abhi{The GR quantities $( f_{\ell m 0}^{GR},\tau_{\ell m 0}^{GR})$ are
constructed using the NR--fitting--formula from Refs.~\cite{Taracchini:2013rva,Hofmann:2016yih}, and are functions of the initial masses and spins, $(m_1, m_2, \chi_1, \chi_2)$. For every single instance of waveform generation, we choose a value for $(m_1, m_2, \chi_1, \chi_2, \delta f_{\ell m 0}, \delta \tau_{\ell m 0})$. While the mass and spin value go into the construction of $( f_{\ell m 0}^{\text{GR}}, \tau _{\ell m 0}^{\text{GR}})$, these GR quantities and then combined with the values of $(\delta f_{\ell m 0}, \delta \tau_{\ell m 0})$ as shown in the Eqs.~\ref{eq:nongr_freqs_a}-~\ref{eq:nongr_freqs_b}.. Hence,
\begin{subequations}
\begin{eqnarray}
f_{\ell m 0} &=& f_{\ell m 0}(m_1, m_2, \chi_1, \chi_2, \delta f_{\ell m 0}, \delta \tau_{\ell m 0})\\ 
\tau _{\ell m 0} &=& \tau _{\ell m 0}(m_1, m_2, \chi_1, \chi_2, \delta f_{\ell m 0}, \delta \tau_{\ell m 0})
\end{eqnarray}
\end{subequations}
We denote such a parameterized waveform model $\pSEOB$~\footnote{This
waveform model is called {\tt pSEOBNRv4HM} in LAL.}.} We note that when leaving $\sigma_{\ell m}$ to vary
freely, the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell
  m}(t)$ in general also differ from the GR predictions, since
those functions depend on the QNM complex frequencies, as can be seen
from the expressions for $c_{i,c}^{\ell m}$ and $d_{1,c}^{\ell m}$ in Eqs.~(\ref{c1}),
(\ref{c2}), and (\ref{d1}). As a consequence, the ringdown signal (amplitude and phase) 
soon after merger deviates from the one predicted by GR.
